---
title: "macro_analysis_NAME"
subtitle: "BIO 131 Project Data Analysis"
author: "Your Name"
date: "March , 2024"
format: 
  html:
    code-overflow: wrap
    embed-resources: true
---

# Getting Setup

Keep all of your analysis in this same .qmd file and keep all the associated files in the project folder. Make sure the project name is showing in the upper right of the RStudio window.

## Load packages

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse) ## for readr, dplyr, ggplot2
library(vegan) ## for calculating diversity measures (you can remove this if you are not calculating diversity measures)
library(ggbeeswarm) ## for jittering points (you can remove this is you did not use a dotplotv)
library(flextable) ## for making formatted tables that can be pasted into Word or Google Docs

```

## Import Data

Figure out which data file(s) you need to import (see the metadata files for more information about each data file).

Find the code you need by clicking on the "script_templates" folder then selecting the "importing_data" folder in the "Files" tab in the lower right area of RStudio.

Copy and paste the code you need into the chunk below (Do **NOT** run the code directly from the script template, copy and paste the template code below)

```{r}
#| label: import data



```

# Data Wrangling

Find the code you need to prepare or "wrangle" your data to make any calculations needed and combine any datasets you need to combine by looking in the "data_wrangling" folder within the "script_templates" folder. Copy and paste the code you need in the chunk below.

Note, once you have completed this code chunk, you should have a new data frame that has ONE value per sampleID. You will then use this new data frame going forward.

```{r}
#| label: data wrangling



```

# Data Description

## Summary Table of Sampling Effort

Make a table of the data that you used in your analysis (how many samples and how much area was sampled in each of the groups) to include in the Data Analysis Methods section of your paper. Go to the "data_description" script templates and select the correct version of the 'sample_table' script based on which factors you are using and paste that code below.

```{r}
#| label: tbl-sample-table
#| tbl-cap: Sample Summary Table
#| include: true





```

## Descriptive Statistics

Use the template code below to calculate the descriptive statistics for your variable of interest.

The output of this code should NOT all be reported in your paper. Decide which (if any) of the descriptive statistics you should report and make sure to round the values.

Descriptive statistics can be reported in the text of the results (see this [example results section](https://docs.google.com/document/d/1vwyOSmhZzTuIfAEbpeiv3yCsU1UMZDsTXgnbE80nPTU/edit?usp=sharing)) if relatively simple, or you can make a formatted table of just the necessary, rounded values to include in your paper.

```{r}
#| label: tbl-descriptive-stats
#| tbl-cap: Descriptive Statistics
#| include: true

#see https://davidgohel.github.io/flextable/ for more information and formatting options

#set the variable you want to summarize
x.var <- "___" #put the name of the column you want to summarize in the blank here

#calculate descriptive stats
df.sum <- my.df |> 
  group_by(___) |> #put the grouping variable(s) here
  filter(!is.na(.data[[x.var]])) |> # remove missing values 
  
  #remove the lines below that you don't need 
  #you won't usually need ALL these values
  #change the digits as needed
  summarise(mean = round(mean(.data[[x.var]]), digits=2), 
            SD = signif(sd(.data[[x.var]]), digits=2),
            median = round(median(.data[[x.var]]), digits=2),
            IQR = round(IQR(.data[[x.var]]), digits=2),
            min = round(min(.data[[x.var]]), digits=2),
            max = round(max(.data[[x.var]]), digits=2),
            N = n())

#create the table
ft <- flextable(df.sum,
                cwidth=0.75) |>  #can vary cell width as needed
  
  #bold the headings
  bold(part="header") |> 
  
  #center columns
  align(align = "center", part = "all" )

#print the table
#right click on the table, choose select all, 
#choose copy, then paste in your document
#finish formatting as needed in your document
ft

```

# Data Visualization

Find the code you need to visualize (=graph) your data by looking in the "graphing" folder within the "script_templates" folder. Copy and paste the code you need in the chunk below.

```{r}
#| label: graph


```

To save your graph with the `ggsave()` function, you need to name the resulting file with surrounding " ", and indicate the size via height, width, and units. Don't forget to save the graph with a dpi call at 300-500 to make it nice and crisp! Look at the `ggsave()` help file for more information and options.

Note that the code below will save the most recent graph you made and put it into the "images" folder (make sure to replace the word "Name" for your name within the filename so you will be able to tell which graph is yours later)

```{r}
#| label: Save your graph
#| eval: false


# save the graph!
ggsave(filename="images/MacroGraph_Name.png",  #recommended to use.png or .jpg file types
       height = 5, width = 8, units = "in", 
       dpi = 300)

```

The above code will save the last graph you made and call it "MacroGraph_Name.png" (make sure you replaced Name with your name. for example: MacroGraph_Pratt.png). This file will appear in the images folder within the project folder under the Files tab in the lower right quadrant of this window.

If you are working on the Smith server, you will first need to export this file to your computer:

-   Check the box next to the "MacroGraph_Name.png" file

-   Go to the "More" menu with the gear icon and click the drop down arrow

-   Select "Export" and then save the file to your computer

You can now insert this graph into your Results document (see the lab notebook entry for today for more details).

# Citing R in Your Paper

In the Data Analysis Methods section of your paper, you need to cite how the data were analyzed including (1) any calculations done (for example, how was density calculated?), and (2) what software was used for the analysis.

Here is a detailed explanation of How to Cite R in your Methods.

You need to cite base R but how will you know what version of R you are using?

You should also cite the **most important packages used**.

All of you should cite the {tidyverse} super package since we used {reader}, {dplyr}, and {ggplot2} to import, wrangle, and graph our data. If you used the dot plot in your paper, then you also used {ggbeeswarm} which is NOT part of the {tidyverse} so it needs to be cited separately. If you used the formatted table in your paper, then cite the {flextable} package. If you used {vegan} to calculate any diversity indices then make sure to cite it.

To properly cite packages, you need to find out the information needed for each package. Run the code below to determine the version of R and packages used. (note that you do NOT need to cite ALL of the packages that the `sessionInfo` function spits out. See above for what you need to cite)

```{r}
#| label: sessionInfo

sessionInfo()

```

The above is the information you need to cite which version of R as well as any packages you used. While the version number is necessary, you also need more information to cite R and the packages. Refer to [How to Cite R in your Methods](https://docs.google.com/presentation/d/1RG4_R-MDOy1vbMz-M1gfOcMtgqWu9Kvv_R4ex5QdcVk/edit?usp=sharing) to help you figure out how to cite in the text as well as the full citation in the Literature Cited section at the end of your paper.

Use the code below to get more necessary information to cite R itself

```{r}
#| label: cite R

#citation info for base R 
citation()

```

Use the code below to cite the {tidyverse} package and then repeat the code replacing "tidyverse" with any other packages you need to cite

```{r}
#| label: cite tidyverse

#citation info for tidyverse package 
citation("tidyverse")

```


## Render this File to HTML & Print to PDF

Don't forget to "render" your .qmd file when you are done.

-   Before you Render

    -   Make sure that each code chunk has an unique label name or no name

    -   Save this file and give it the name "macro_analysis_Name.qmd" (replace your name with the "Name" part)

-   Render this Quarto file to HTML (just select the Render menu item at the top of this window)

-   If the HTML file didn't already open in your browser, then click on the name of the file in the Files window and select "View in Web Browser"

-   Once the HTML file is open in your web browser, then use the print function within your browser to save it as a PDF. Please name your PDF file "macro_analysis_Name.pdf" (replace your name with the "Name" part)

-   Make sure the PDF file is saved somewhere on your computer that you can get to

-   Place a copy of the PDF file in your GROUP Google Drive folder for this class

Render straight to HTML, open the HTML file in your browser, and then "print" to a PDF file.

Make sure you add your name to the PDF file name: macro_analysis_Pratt.pdf

Save a copy the PDF in your GROUP FOLDER.
